Peaks Over Threshold for Bursty Time Series

### 2018-04-23 article_springer/article_springer.Rmd
model is now called CTRE + minor changes

  This article generalizes the Peaks Over Threshold (POT) model to the setting 
  where inter-event times are heavy-tailed. For high thresholds and 
  infinite-mean waiting times, we show that the times between threshold 
  crossings are Mittag-Leffler distributed, and thus form a "fractional 
  Poisson Process" which generalizes the standard Poisson Process.
  We provide graphical means of estimating model parameters and assessing 
  model fit. Along the way, we apply our inference method to a real-world
  bursty time series, and 
  show how the Mittag-Leffler distribution has a memory which affects 
  the predictive distribution of the next extreme event.

@@

We review the marked renewal process in Section 2, and 
derive a scaling limit theorem for inter-exceedance times in Section 3.
We give a statistical procedure to estimate model parameters via stability plots 
in Section 5, but to set the stage we first discuss inference for the 
Mittag-Leffler distribution in Section 4. 
Diagnostic plots for model criticism are discussed in Section 6. 
In Section 7, we discuss the non-memorylessness property of the 
Mittag-Leffler distribution, and how it affects the predictive distribution 
for the next threshold crossing event. 
Section 8 concludes.
For all statistical computations we have used R [@R]. 
Code used for the analysis in this article has been organized into 
an R package `CTRM` (<https://github.com/UNSW-MATH/CTRM>).
The source code for this manuscript is available online at <https://github.com/UNSW-MATH/bursty-POT>.

# Marked Renewal Process (MRP)

@@

: Let $(W,J), (W_1, J_1), (W_2, J_2), \ldots$ be i.i.d. pairs of random 
variables, where the $W_k > 0$ are interpreted as the *waiting times* or 
inter-arrival times and $J_k \in [x_L, x_R]$ as the *event magnitudes* 
or marks
($x_L \in [-\infty, +\infty), x_R \in (-\infty, +\infty]$). 
If $W$ and $J$ are independent, the MRP is said to be *uncoupled*.

In this setup, the $k$-th magnitude $J_k$ occurs at time
$T_k = W_1 + \ldots + W_k$. 
Based on this simple model, we derive the following two random variables:

**Definition:**

: Given a threshold $\ell \in (x_L, x_R)$, 
consider the stopping time 
$$\tau(\ell) := \min\{k: J_k > \ell\},\quad \ell \in (x_L, x_R).$$
The random variables
$$X(\ell) = J_{\tau(\ell)} - \ell, \quad 
T(\ell) = \sum_{k=1}^{\tau(\ell)} W_k$$
are called the *exceedance* resp. *time until exceedance* of level $\ell$.

By restarting the MRP at $\tau(\ell)$, one can inductively construct the
two i.i.d. sequences $T(\ell,n)$ and $X(\ell, n)$, $n \in \mathbb N$.
Moreover, since we assume the uncoupled case, the two sequences are independent.[^1]

@@

```{r thresholdedBursty, message=FALSE, fig.height=7, fig.width=9, fig.cap="\\label{fig:thresholdedBursty}Exceedances (red) and Times until Exceedance (durations between blue crosses) for a given threshold $\\ell$ (dashed line)."}
#rpareto <- function(n,tail){
#          if(tail<=0) stop("tail must be positive")
#          rp <- (runif(n)/gamma(1-tail))^(-1/tail)
#          rp}
tail <- params$tail
n <- params$n 
sigma <- (cos(pi*tail/2))^(1/tail)
times <- cumsum(stabledist::rstable(n, tail, 1, sigma, pm=1))
#times<-cumsum(rpareto(n,tail))
magnitudes <- extRemes::revd(n, scale = 1, shape = 0)
sim_ctrm <- ctrm(data.frame(times, magnitudes)) 
par(mfrow = c(2,1))
plot(sim_ctrm, p = 0.01, main = "Simulated MRP") 
flares %>% ctrm() %>% plot(p = 0.02, log = 'y', main = "HXRBS data") 
```

Figure \ref{fig:thresholdedBursty} shows a simulated dataset in the top 
panel, where $W$ has a stable distribution with tail parameter 
$\beta =$ `r params$tail` (and skewness $1$ and location $0$), and 
where $J$ is from a standard Gumbel distribution. In the bottom panel, 
we plot a time series of solar flare intensities derived from a NASA
dataset [@HXRBS][^*]. Clearly, the simulated data exhibit long intervals 
_without any_ events, whereas in the real-world dataset events appear 
continuously. The threshold exceedances, however, appear to have similar
statistical behaviour in both models. Observations below a threshold are 
commonly discarded in Extreme Value Theory; likewise, our MRP model
interprets these observations as noise and filters them out.

@@

: Let the waiting times $J_k$ be in the domain of attraction of a 
positively skewed sum-stable law with stability parameter $0 < \beta < 1$; 
more precisely,
\begin{align} \label{eq:stability}
\frac{W_1 + \ldots + W_n}{b(n)} \overset{d}{\longrightarrow} D, 
\quad n \to \infty
\end{align}
for a function $b(n)$ which is regularly varying at $\infty$ with 
parameter $1/\beta$, and where $\mathbf E[\exp(-sD)] = \exp(-s^\beta)$. 
Write $p := \mathbf P(J > \ell)$. Then the weak convergence
$$
\frac{T(\ell)} {b(1/p)} \to W_\beta \quad \text{ as } \quad \ell \uparrow x_R
$$
holds, where the Mittag-Leffler random variable $W_\beta$ is defined on 
the positive real numbers via 
$$
\mathbf E[\exp(-sW_\beta)] = \frac{1}{1+s^\beta}.
$$

\noindent For a scale parameter $\sigma > 0$, we write 
${\rm ML}(\beta, \sigma)$ for the distribution of $\sigma W_\beta$. 
The Mittag-Leffler distribution with parameter $\beta \in (0,1]$ is 
a heavy-tailed positive distribution for $\beta < 1$, with infinite mean. 
However, as $\beta \uparrow 1$, ${\rm ML}(\beta, \sigma)$ converges 
weakly to the exponential distribution ${\rm Exp}(\sigma)$.
This means that although its moments are all infinite, the Mittag-Leffler
distribution may (if $\beta$ is close to 1) be indistinguishable from the 
exponential distribution, from the point of view of applied statistics. 
For a detailed reference on the Mittag-Leffler distribution, see e.g. 
@Haubold11, and for algorithms, see e.g. the R package `MittagLeffleR`
[@MittagLeffleR].

@@

flare_mags <- flares %>% ctrm() %>% thin(k=500) %>% coredata()
flares_exponent <- flare_mags %>% qqestplot_static()

@@

However, as described nicely by @Resnick97, the less 
similar a heavy-tailed distribution is to a Pareto distribution, the less
useful a Hill or QQ-Plot estimator becomes. 
The bottom-left panel of Figure \ref{fig:QQ-plots}, for instance, shows the 
static QQ-estimator plot for $2000$ draws from the Mittag-Leffler distribution with 
tail parameter $0.8$. The stretched exponential shape means 
that the Mittag-Leffler distribution has more probability near $0$ than 
the Pareto distribution, severely biasing the estimator downwards 
(`r round(ML_exponent, 2)`). 
Even by looking at different cutoffs via the dynamic QQ-estimator plot 
(bottom-center panel) one hardly identifies $0.8$ as a stable value, and 
thus as a clear candidate for a tail parameter estimate.

QQ-estimators, and the closely related Hill estimator, are hence not 
suitable to detect a heavy-tailed Mittag-Leffler distribution. Indeed, 
QQ-estimator plots of exponentially distributed data are virtually 
indistinguishable from the bottom-left panel. This shows that 
_Mittag-Leffler distributed data may look like exponentially distributed 
data if examined via a QQ-estimator_.

@@

Hence if there is some prior expectation that the data are drawn from the 
Mittag-Leffler distribution
(as is the the case for threshold exceedance times),
then we recommend avoiding QQ-estimators and Hill plots altogether, 
and instead examining QQ-plots directly 
(see Figure \ref{fig:QQ-plots}, right column). 
The scale parameter $\sigma$ is irrelevant for a QQ-Plot. 
The tail parameter $\beta$ can be estimated quickly via the log-moment 
method by @Cahoy2013, or via maximum likelihood. 
Both estimators are implemented in `MittagLeffleR` 
[@MittagLeffleR]. 
Since the exponential distribution is nested in the Mittag-Leffler 
family of distributions, a standard likelihood ratio test can be 
performed, with the exponential distribution as a null model against 
a Mittag-Leffler distribution as the alternative. 
As an example, the threshold crossing times for the solar flare dataset
yield a difference in deviance of 
$\approx$ `r round(2 * (mlmle_out$loglik - mlmle_out$null_loglik))`, 
which evaluates to a strongly significant $\chi^2_1$ test statistic 
(p-value = $10^{`r round(log10(mlmle_out$p_value))`}$) for the null
hypothesis $\beta = 1$.

@@

The Theorem in Section \ref{sec:scaling} implies that for high 
threshold $\ell$ we may 
approximate the distribution of $T(\ell)$ with an 
${\rm ML}(\beta, b(1/p))$ distribution, 
where $b(\,)$ varies regularly at $\infty$ with parameter $1/\beta$.
Building on the POT (Peaks Over Threshold) method, we propose the 
following estimation procedure for the distribution of $T(\ell)$:

@@

```{r flare-diagnostics-1, fig.height=7, fig.cap="Diagnostic plots for the solar flare data: auto-correlation function.\\label{fig:flare-diagnostics-1}"}
flares_ctrm <- flares %>% ctrm() %>% thin(k = 200)
acf(flares_ctrm)
```

@@

The MRP model is based on three main assumptions, which are repeated 
below. For each assumption, we suggest one means of checking if it holds:

@@

Figures \ref{fig:flare-diagnostics-1} and \ref{fig:flare-diagnostics-2} 
show the diagnostic plots for a 
minimum threshold chosen at the 200th order statistic. 
There is some residual autocorrelation for the sequence of threshold 
exceedance times that is not accounted for by the MRP model. 
The fit with a Mittag-Leffler distribution ($\beta = 0.8$) is good, 
though there are signs that the power-law tail tapers off for very large 
inter-threshold crossing times. 
There is no apparent dependence between threshold exceedance times and 
event magnitudes seen in the copula plot.

@@

When we introduced the MRP model, we assumed that all events are i.i.d. 
This assumption is likely sufficient but not necessary for our limit theorem to hold. 
Morevoer, any data below a (minimum) threshold $\ell_0$ is discarded in 
our inference procedure, and hence need not satisfy the i.i.d. assumption.
For the purposes of statistical inference, we merely require that the 
inter-threshold-crossing times are i.i.d.

For some bursty time series, the current MRP model may be too rigid, 
because often the heavy-tailed character of the inter-arrival time
distribution does not hold at all time scales; rather, it often applies at 
short and intermediate time scales, and is truncated or tempered 
(reverting to an exponential distribution) at 
very long time scales, see e.g. @MeerschaertRoyQin and @Aban06. 
In such situations, a "tempered" Mittag-Leffler distribution may provide 
a more realistic fit, which we aim to introduce in follow-up work.

>>>>

  This article generalizes the Peaks Over Threshold (POT) model to the setting 
  where inter-event times are heavy-tailed. For high thresholds and 
  infinite-mean waiting times, we show that the times between threshold 
  crossings are Mittag-Leffler distributed, and thus form a "fractional 
  Poisson Process" which generalizes the standard Poisson Process.
  We provide graphical means of estimating model parameters and assessing 
  model fit. Along the way, we apply our inference method to a real-world
  bursty time series, and 
  show how the memory of the Mittag-Leffler distribution affects 
  the predictive distribution for the time until the next extreme event.

@@

We review the marked renewal process in Section 2, and 
derive a scaling limit theorem for inter-exceedance times in Section 3.
We give a statistical procedure to estimate model parameters via stability plots 
in Section 5, but to set the stage we first discuss inference for the 
Mittag-Leffler distribution in Section 4. 
Diagnostic plots for model criticism are discussed in Section 6. 
In Section 7, we discuss the non-memorylessness property of the 
Mittag-Leffler distribution, and how it affects the predictive distribution 
for the next threshold crossing event. 
Section 8 concludes.
For all statistical computations we have used R [@R]. 
Code used for the analysis in this article has been organized into 
an R package `CTRM` (<https://github.com/UNSW-MATH/CTRM>).
The source code for the figures generated in this manuscript is 
available online at <https://github.com/UNSW-MATH/bursty-POT>.

# Continuous Time Random Exceedances (CTRE)

As a model for extreme observations, we use a Marked Renewal Process (MRP):

@@

: Let $(W,J), (W_1, J_1), (W_2, J_2), \ldots$ be i.i.d. pairs of random 
variables, where the $W_k > 0$ are interpreted as the *waiting times* 
and $J_k \in [x_L, x_R]$ as the *event magnitudes* 
($x_L \in [-\infty, +\infty), x_R \in (-\infty, +\infty]$). 
If $W$ and $J$ are independent, the Marked Renewal Process is said to be *uncoupled*. 
\qed

Note that the $k$-th magnitude $J_k$ occurs at time
$T_k = W_1 + \ldots + W_k$. 
Based on an MRP, we define the Continuous Time Random Exceedance model
(CTRE) as follows:

**Definition (CTRE):**

: Given a threshold $\ell \in (x_L, x_R)$, 
consider the stopping time 
$$\tau(\ell) := \min\{k: J_k > \ell\},\quad \ell \in (x_L, x_R).$$
Define the pair of random variables $(X(\ell), T(\ell))$ via 
$$X(\ell) = J_{\tau(\ell)} - \ell, \quad 
T(\ell) = \sum_{k=1}^{\tau(\ell)} W_k.$$
By restarting the MRP at $\tau(\ell)$, inductively define the
two i.i.d. sequences $T(\ell,n)$ and $X(\ell, n)$, $n \in \mathbb N$, 
called the "interarrival times" and the "exceedances", respectively.
The pair sequence $(T(\ell, n), W(\ell, n))_{n \in \mathbb N}$ is called 
a Continuous Time Random Exceedance model (CTRE). 
If the underlying MRP is uncoupled, then the CTRE is also called 
uncoupled. 
\qed

```{r thresholdedBursty, message=FALSE, fig.height=7, fig.width=9, fig.cap="\\label{fig:thresholdedBursty}Exceedances (red) and Times until Exceedance (durations between blue crosses) for a given threshold $\\ell$ (dashed line)."}
tail <- params$tail
n <- params$n 
sigma <- (cos(pi*tail/2))^(1/tail)
times <- cumsum(stabledist::rstable(n, tail, 1, sigma, pm=1))
#times<-cumsum(rpareto(n,tail))
magnitudes <- extRemes::revd(n, scale = 1, shape = 0)
sim_ctrm <- ctrm(data.frame(times, magnitudes)) 
par(mfrow = c(2,1))
plot(sim_ctrm, p = 0.01, main = "Simulated MRP") 
flares %>% ctrm() %>% plot(p = 0.02, log = 'y', main = "HXRBS data") 
```

In this article, we restrict ourselves to the uncoupled case, where
the two sequences $X(\ell, n)_{n \in \mathbb N}$
and $T(\ell, n)_{n \in \mathbb N}$ are independent.[^1]
Figure \ref{fig:thresholdedBursty} shows a simulated dataset in the top 
panel, where $W$ has a stable distribution with tail parameter 
$\beta =$ `r params$tail` (and skewness $1$ and location $0$), and 
where $J$ is from a standard Gumbel distribution. In the bottom panel, 
we plot a time series of solar flare intensities derived from a NASA
dataset [@HXRBS][^*]. Clearly, the simulated data exhibit long intervals 
_without any_ events, whereas in the real-world dataset events appear 
continuously. The threshold exceedances, however, appear to have similar
statistical behaviour in both models. Observations below a threshold are 
commonly discarded in Extreme Value Theory; likewise, the CTRE model
interprets these observations as noise and filters them out.

@@

@@

: Let the waiting times $J_k$ be in the domain of attraction of a 
positively skewed sum-stable law with stability parameter $0 < \beta < 1$; 
more precisely,
\begin{align} \label{eq:stability}
\frac{W_1 + \ldots + W_n}{b(n)} \overset{d}{\longrightarrow} D, 
\quad n \to \infty
\end{align}
for a function $b(n)$ which is regularly varying at $\infty$ with 
parameter $1/\beta$, and where $\mathbf E[\exp(-sD)] = \exp(-s^\beta)$. 
Write $p := \mathbf P(J > \ell)$. Then the weak convergence
$$
\frac{T(\ell)} {b(1/p)} \to W_\beta \quad \text{ as } \quad \ell \uparrow x_R
$$
holds, where the Mittag-Leffler random variable $W_\beta$ is defined on 
the positive real numbers via 
$$
\mathbf E[\exp(-sW_\beta)] = \frac{1}{1+s^\beta}.
$$
\qed

\noindent For a scale parameter $\sigma > 0$, we write 
${\rm ML}(\beta, \sigma)$ for the distribution of $\sigma W_\beta$. 
The Mittag-Leffler distribution with parameter $\beta \in (0,1]$ is 
a heavy-tailed positive distribution for $\beta < 1$, with infinite mean. 
However, as $\beta \uparrow 1$, ${\rm ML}(\beta, \sigma)$ converges 
weakly to the exponential distribution ${\rm Exp}(\sigma)$.
This means that although its moments are all infinite, the Mittag-Leffler
distribution may (if $\beta$ is close to 1) be indistinguishable from the 
exponential distribution, for the purposes of applied statistics. 
For a detailed reference on the Mittag-Leffler distribution, see e.g. 
@Haubold11, and for algorithms, see e.g. the R package `MittagLeffleR`
[@MittagLeffleR].

@@

flare_mags <- flares %>% ctrm() %>% thin(k=500) %>% magnitudes()
flares_exponent <- flare_mags %>% qqestplot_static()

@@

However, as described nicely by @Resnick97, the less 
similar a heavy-tailed distribution is to a Pareto distribution, the less
useful a Hill or QQ-Plot estimator becomes. 
The bottom-left panel of Figure \ref{fig:QQ-plots}, for instance, shows the 
static QQ-estimator plot for $2000$ draws from the Mittag-Leffler distribution with 
tail parameter $0.8$. The stretched exponential shape means 
that the Mittag-Leffler distribution has more probability near $0$ than 
the Pareto distribution, severely biasing the estimator downwards 
(`r round(ML_exponent, 2)`). 
Even by looking at different cutoffs via the dynamic QQ-estimator plot 
(bottom-center panel) one hardly identifies $0.8$ as a clear candidate for 
a tail parameter estimate.

QQ-estimators, and the closely related Hill estimator, are hence not 
suitable to detect a heavy-tailed Mittag-Leffler distribution. Moreover, 
QQ-estimator plots of exponentially distributed data are virtually 
indistinguishable from the bottom-left panel. This shows that 
_Mittag-Leffler distributed data may look like exponentially distributed 
data if examined via a QQ-estimator_.

@@

Hence if there is some prior expectation that the data are drawn from the 
Mittag-Leffler distribution
(as is the the case for threshold exceedance times),
then we recommend avoiding QQ-estimators and Hill plots altogether, 
and instead examining QQ-plots directly, on a logarithmic scale 
(see Figure \ref{fig:QQ-plots}, right column).
The scale parameter $\sigma$ is irrelevant for a QQ-Plot. 
The tail parameter $\beta$ can be estimated quickly via the log-moment 
method by @Cahoy2013, or via maximum likelihood. 
Both estimators are implemented in `MittagLeffleR` 
[@MittagLeffleR]. 
Since the exponential distribution is nested in the Mittag-Leffler 
family of distributions, a standard likelihood ratio test can be 
performed, with the exponential distribution as a null model against 
a Mittag-Leffler distribution as the alternative. 
As an example, the threshold crossing times for the solar flare dataset
(Figure \ref{fig:flare-diagnostics-2}, right panel)
yield a difference in deviance of 
$\approx$ `r round(2 * (mlmle_out$loglik - mlmle_out$null_loglik))`, 
which evaluates to a strongly significant $\chi^2_1$ test statistic 
(p-value = $10^{`r round(log10(mlmle_out$p_value))`}$) for the null
hypothesis $\beta = 1$.

@@

The Theorem in Section \ref{sec:scaling} implies that for a high 
threshold $\ell$ we may 
approximate the distribution of $T(\ell)$ with an 
${\rm ML}(\beta, b(1/p))$ distribution, 
where $b(\,)$ varies regularly at $\infty$ with parameter $1/\beta$.
Building on the POT (Peaks Over Threshold) method, we propose the 
following estimation procedure for the distribution of $T(\ell)$:

@@

```{r flare-diagnostics-1, fig.height=7, fig.cap="Diagnostic plots for the solar flare data: auto-correlation function.\\label{fig:flare-diagnostics-1}"}
flares_ctrm <- flares %>% ctrm() %>% thin(k = 150)
acf(flares_ctrm)
```

@@

The CTRE model is based on three main assumptions, which are repeated 
below. For each assumption, we suggest one means of checking if it holds:

@@

Figures \ref{fig:flare-diagnostics-1} and \ref{fig:flare-diagnostics-2} 
show the diagnostic plots for a 
minimum threshold chosen at the 200th order statistic. 
There is some residual autocorrelation for the sequence of threshold 
exceedance times that is not accounted for by the CTRE model. 
The fit with a Mittag-Leffler distribution ($\beta = 0.8$) is good, 
though there are signs that the power-law tail tapers off for very large 
inter-threshold crossing times. 
There is no apparent dependence between threshold exceedance times and 
event magnitudes seen in the copula plot.

@@

When we introduced the CTRE model, we assumed that all events are i.i.d. 
This assumption is likely sufficient but not necessary for our limit theorem to hold. 
Morevoer, any data below a (minimum) threshold $\ell_0$ is discarded in 
our inference procedure, and hence need not satisfy the i.i.d. assumption.
For the purposes of statistical inference, we merely require that the 
inter-threshold-crossing times are i.i.d.

For some bursty time series, the current CTRE model may be too rigid, 
because often the heavy-tailed character of the inter-arrival time
distribution does not hold at all time scales; rather, it often applies at 
short and intermediate time scales, and is truncated or tempered 
(reverting to an exponential distribution) at 
very long time scales, see e.g. @MeerschaertRoyQin and @Aban06. 
In such situations, a "tempered" Mittag-Leffler distribution may provide 
a more realistic fit, which we aim to introduce in follow-up work.
